NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PS lipids

### 2018-01-20 Manuscript/manuscriptPS.tex
Update manuscript

@@

PC lipid headgroup order parameters can used to measure ion binding
affinity, because their magnitude is proportional
to the amount of bound charge in bilayer \cite{seelig87,catte16}.
The molecular electrometer concept can be used also
for bilayers containing PC lipids mixed with charged lipids \cite{borle85,macdonald87,roux90}.
This is demonstrated in Fig. %s \ref{OrderParametersWithCaCl},
%\ref{OrderParametersWithCaClBELOW1M} and
\ref{OrderParameterCHANGESWithCaClBELOW1M},
showing the changes of PC headgroup order parameters %for PC headgroup $\alpha$ and $\beta$ carbons
as a function of CaCl$_2$ concentration in the presence of different amounts of
negatively charged PS or PG lipids.
%\begin{figure}[]
%  \centering
%  \includegraphics[width=9.0cm]{../Figs/LIPIDSwithCaCl.eps}
%  \caption{\label{OrderParametersWithCaCl}
%    PC headgroup order parameters as a function of CaCl concentration from experiments containing charged lipids.
%    Pure DPPC data from \cite{akutsu81}, pure POPC data from \cite{altenbach84}, 
%    POPC:POPS mixture data from \cite{roux90}, POPC:POPG mixture data with 0.1M NaCl from \cite{macdonald87}
%    and POPC:POPG mixture data without NaCl from \cite{borle85}.
%  }
%  \todo{Check the NaCl concentrations in the samples.}
%\end{figure}
%\begin{figure}[]
%  \centering
%  \includegraphics[width=9.0cm]{../Figs/LIPIDSwithCaClBELOW1M.eps}
%  \caption{\label{OrderParametersWithCaClBELOW1M}
%    Figure \ref{OrderParametersWithCaCl} zoomed to smaller concentrations.
%  }
%\end{figure}
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/CHANGESwithCaCl.eps}
  \caption{\label{OrderParameterCHANGESWithCaClBELOW1M}
    The change of PC headgroup order parameters
    in the presence of different amount of negatively charged lipids
    respect to the values without added CaCl$_2$.
    The original data is the same as in Fig. \ref{OrderParametersWithCaCl}. 
  }
\end{figure}
%PC headgroup order parameters increase when negatively charged
%PS or PG are added to PC bilayer in the absense of added CaCl$_2$,
%as expected based on electrometer concept \cite{seelig87}
%(see Fig. \ref{OrderParametersWithCaClBELOW1M}).
%In electrometer concept this is explained by the tilting of
%headgroup more parallel to membrane normal \cite{??}.
The decrease of order parameters with CaCl$_2$
is more pronounced for systems with more
negatively charged lipids.% (see Fig. \ref{OrderParameterCHANGESWithCaClBELOW1M}).
Order parameters reach the values of pure PC bilayer close to CaCl$_2$ concentrations of $\sim$ 50-300mM.
At this point the Ca2+ binding presumably fully cancels the charge from negative lipids and
overcharging occurs above these concenterations.
The interpretation of this data and some other results has been that \cite{seelig90}
\begin{displayquote}
  {\it ''(i) Ca$^{2+}$ binds to neutral lipids (phosphatidylcholine, phosphatidylethanolamine) and negatively charged lipids
    (phosphatidylglycerol) with approximately the same binding constant of K = 10-20 M$^{-1}$; \\
    (ii) the free Ca$^{2+}$
    concentration at the membrane interface is distinctly enhanced if the membrane carries a negative surface
    charge, either due to protein or to lipid; \\
    (iii) increased inter-facial Ca$^{2+}$ also means increased amounts
    of bound Ca$^{2+}$ at neutral and charged lipids; \\
    (iv) the actual binding step can be described by a Langmuir
    adsorption isotherm with a 1 lipid:1 Ca$^{2+}$ stoichiometry, provided the interfacial concentration C$_M$, is
    used to describe the chemical binding equilibrium.''}
\end{displayquote}

@@

>>>>

PC lipid headgroup order parameters can used to measure ion binding
affinity, because their magnitude is proportional
to the amount of bound charge in bilayer \cite{seelig87,catte16}.
The molecular electrometer concept can be used also
for bilayers containing PC lipids mixed with charged lipids \cite{borle85,macdonald87,roux90}.
This is demonstrated in Fig. %s \ref{OrderParametersWithCaCl},
%\ref{OrderParametersWithCaClBELOW1M} and
\ref{OrderParameterCHANGESWithCaClBELOW1M},
showing the changes of PC headgroup order parameters %for PC headgroup $\alpha$ and $\beta$ carbons
as a function of CaCl$_2$ concentration in the presence of different amounts of
negatively charged PS or PG lipids.

%PC headgroup order parameters increase when negatively charged
%PS or PG are added to PC bilayer in the absense of added CaCl$_2$,
%as expected based on electrometer concept \cite{seelig87}
%(see Fig. \ref{OrderParametersWithCaClBELOW1M}).
%In electrometer concept this is explained by the tilting of
%headgroup more parallel to membrane normal \cite{??}.
The decrease of order parameters with CaCl$_2$
is more pronounced for systems with more
negatively charged lipids.% (see Fig. \ref{OrderParameterCHANGESWithCaClBELOW1M}).
Order parameters reach the values of pure PC bilayer close to CaCl$_2$ concentrations of $\sim$ 50-300mM.
At this point the Ca2+ binding presumably fully cancels the charge from negative lipids and
overcharging occurs above these concenterations.
The interpretation of this data and some other results has been that \cite{seelig90}
\begin{displayquote}
  {\it ''(i) Ca$^{2+}$ binds to neutral lipids (phosphatidylcholine, phosphatidylethanolamine) and negatively charged lipids
    (phosphatidylglycerol) with approximately the same binding constant of K = 10-20 M$^{-1}$; \\
    (ii) the free Ca$^{2+}$
    concentration at the membrane interface is distinctly enhanced if the membrane carries a negative surface
    charge, either due to protein or to lipid; \\
    (iii) increased inter-facial Ca$^{2+}$ also means increased amounts
    of bound Ca$^{2+}$ at neutral and charged lipids; \\
    (iv) the actual binding step can be described by a Langmuir
    adsorption isotherm with a 1 lipid:1 Ca$^{2+}$ stoichiometry, provided the interfacial concentration C$_M$, is
    used to describe the chemical binding equilibrium.''}
\end{displayquote}

@@

@@

\section{Cation binding affinity to lipid bilayers with different amount of charge}

%\begin{figure}[]
%  \centering
%  \includegraphics[width=9.0cm]{../Figs/LIPIDSwithCaCl.eps}
%  \caption{\label{OrderParametersWithCaCl}
%    PC headgroup order parameters as a function of CaCl concentration from experiments containing charged lipids.
%    Pure DPPC data from \cite{akutsu81}, pure POPC data from \cite{altenbach84}, 
%    POPC:POPS mixture data from \cite{roux90}, POPC:POPG mixture data with 0.1M NaCl from \cite{macdonald87}
%    and POPC:POPG mixture data without NaCl from \cite{borle85}.
%  }
%  \todo{Check the NaCl concentrations in the samples.}
%\end{figure}
%\begin{figure}[]
%  \centering
%  \includegraphics[width=9.0cm]{../Figs/LIPIDSwithCaClBELOW1M.eps}
%  \caption{\label{OrderParametersWithCaClBELOW1M}
%    Figure \ref{OrderParametersWithCaCl} zoomed to smaller concentrations.
%  }
%\end{figure}
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/CHANGESwithCaCl.eps}
  \caption{\label{OrderParameterCHANGESWithCaClBELOW1M}
    The change of PC headgroup order parameters
    in the presence of different amount of negatively charged lipids
    respect to the values without added CaCl$_2$.
    The original data is the same as in Fig. \ref{OrderParametersWithCaCl}. 
    Pure DPPC data from \cite{akutsu81}, pure POPC data from \cite{altenbach84}, 
    POPC:POPS mixture data from \cite{roux90}, POPC:POPG mixture data with 0.1M NaCl from \cite{macdonald87}
    and POPC:POPG mixture data without NaCl from \cite{borle85}.
  }
\end{figure}
